---
title: "Lattice"
author: "Ryan Yancey"
date: "1/7/2021"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lattice Plotting System

Implemented through `lattice` and `grid` packages.

* `lattice` contains functions like `xyplot` `bwplot` and `levelplot`
  + Unlike `base`, plots are made in a single function call.


* `grid` is what `lattice` builds on top of. (It is rarely called independently).


## Lattice Functions

`xyplot` -- Scatterplots

`bwplot` -- Box-and-whisker plots (box plot)

`histogram` -- Histogram

`stripplot` -- Similar to boxplot, but with points

`dotplot` -- "Violin strings"

`splom` -- Scatterplot matrix

`levelplot`/`contourplot` -- Plots image data

****

**Lattice functions typically have a formula for their first argument.**

`xyplot(y ~ x | f * g, data)`

* y-axis variable, x-axis variable

* f and g are "conditioning" variables (optional)

  + They are categorical variables that will produce multiple plots

## Simple Lattice Plot

```{r plot, echo = TRUE}
library(datasets)
library(lattice)

xyplot(Ozone ~ Wind, data = airquality)
```

## Another Lattice Plot

```{r plot2, echo = TRUE}
airquality <- transform(airquality, Month = factor(Month, labels = c("May", "June", "July", "Aug", "Sept")))

xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5,1), main = "Ozone vs Wind in NY (1973)",
       type = c("p","r"))
```

## Making fake data
```{r makedata, echo = TRUE}
set.seed(10)
x <- rnorm(5000)
f <- rep(0:4, each = 1000)
y <- x + f - f * x + rnorm(5000, sd = 0.2)
f <- factor(f, labels = c("Group 1", "Group 2", "Group 3", "Group 4", "Group 5"))
```

## Using fake data  in panel plot
```{r xyplot, echo = TRUE}
xyplot(y ~ x | f, layout = c(5,1))
```

## Lattice with custom panel functions
```{r xyplot2, echo = TRUE}
xyplot(y ~ x | f, layout = c(5,1), panel = function(x, y, ...){
        panel.xyplot(x, y, ...)
        panel.lmline(x, y, col = 2, lwd = 2)
        panel.abline(h = mean(y), lty = 2)
})
```





